Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-04-24 ms/spaceness.tex
pre-PR edits

%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
%Impacts of Isolation by Distance on Demographic Modeling and Association Studies
%Space is the Place: How Dispersal and Sampling Shape Inference From Genetic Variation in Continuous Space
%Evolution on the Final Frontier: How Dispersal and Sampling Shape Inference From Genetic Data in Continuous Space
\title{Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations. As a result many methods of analyzing genetic data assume that samples are a random draw from a random-mating population, but are applied to real populations that are structured over space by limited dispersal. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and association studies. We find that estimates of population size trajectories and many common summary statistics vary significantly with dispersal, with most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered. In association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome, and this bias is only partially corrected by regressing out principal components positions in the analysis. Last we describe the genealogical process leading to spatial effects we observed on simulated gentoypes, summarize estimates of neighborhood size from empirical studies across a range of organisms, and discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

\section{Introduction}
The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models. Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models describing it are flawed approximations of the underlying process \citep{Felsenstein1975}.
As a result many studies focus on describing geographic structure as a set of random-mating populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{structure} \citep{Pritchard2000} as random-mating units.

Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a random mating population. If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will represent a biased sample of the genetic variation present in the whole population. All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.

Our understanding of how many common summary statistics respond to variation in dispersal distance is limited. Tajima's $D$, for example, is often interpreted as reflecting evidence of selection or past population size changes (CITE \cjb{CJ: what's a good citation here?}). However this statistic is essentially a summary of the site frequency spectrum, which itself reflects variation in branch lengths of the underlying genealogy of sampled individuals. Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these genealogical summaries may also reflect variation in the scale of spatial interactions both over time and across populations. Similarly, the distribution of haplotype block lengths (i.e. stretches of identical sequence) among individuals is sometimes interpreted as reflecting historic demography \citep{Harris2013} or evidence of selection \citep{Garud2015}. However as this distribution is also an indirect estimate of the branch lengths of the sample genealogy it is also sensitive to variation in dispersal and mate selection \citep{Ringbauer2017,Baharian2016}.

@@

The issue of fine-scale structure for putatively "within-population" analyses is particularly important for studies of polygenic selection. This is because the genome-wide allele frequency variation caused by isolation by distance is similar to that expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}. When population structure and the environment covary across space genome-wide association studies (GWAS) may incorrectly infer phenotypic association at many SNPs that have differentiated by drift \citep{Mathieson2012}. Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient. However we have little quantitative idea of the population and environmental parameters that can be expected to lead to bias in GWAS. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed under discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum.

Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.

@@

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, we  also implemented a second version of the SLiM simulation in which mates are selected at random and offspring disperse to a random location on the landscape. We refer to these two models as the "spatial" and "random mating" models. For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}. \ak{add a link to the github repo with code}

@@

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples. The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function \ak{point to code on github}. We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

@@

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.

In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 110 and standard deviation 10.
\begin{equation}
    P \sim \mathcal{N}(110,10)
\end{equation}
Next we simulated clinal environmental influences on phenotype by drawing the phenotypes from independent normal distributions in which the mean was scaled by an individual's x position such that it varied by 2 standard deviations across the map. If $x$ is an individual's x coordinate on a 50x50 landscape, the phenotype is then:
\begin{equation}
    P \sim \mathcal{N}(110+\frac{2x}{5},10)
\end{equation}
Third, we approximate concentrated environmental effects by drawing phenotypes for individuals with x and y coordinates below 20 from a normal distribution with mean 2 standard deviations above the rest of the map. 
\begin{equation}
    P\sim
\begin{cases}
    \mathcal{N}(130,10),& \text{if } x< 20 \text{ and } y<20\\
    \mathcal{N}(110,10),& \text{otherwise}\\
\end{cases}
\end{equation}

@@

\begin{equation}
    P\sim
\begin{cases}
    \mathcal{N}(130,10),& \text{if } d < 3\\
    \mathcal{N}(110,10),& \text{otherwise}\\
\end{cases}
\end{equation}
where $d$ is the distance to the nearest patch.

@@

\subsection{Effects of Dispersal}
Limited dispersal inflates effective population size, creates correlations between genetic and spatial distances, and introduces subtle biases in the site frequency spectrum that are reflected in a positive Tajima's $D$ (Figure \ref{fig:sumstats}). At the extreme low end of dispersal distance this can result in an up to three-fold increase in genetic diversity relative to random-mating expectations. These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g. inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling). This general pattern is similar to that predicted by the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.

@@

\subsection{Effects of Sampling}
One of the most important differences between random mating and spatial models is the effect of sampling: in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. As expected we found that random sampling provides the most accurate summary of genetic diversity across the landscape. However this strategy is often impractical for empirical studies. In reality the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site means that most studies follow something closer to the "point" sampling strategy we simulated, in which multiple individuals are sampled from nearby points on the landscape. For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. In classical studies of Drosophila variation the situation is considerably worse, in which a single orchard might be sampled with bated traps for instance.

@@

\subsection{Demography}
Classical population genetic models collapse many elements of life history variation into a single parameter, $N_{e}$, which is then taken to reflect the degree of variation present in the population when modeling the effects of selection or migration. Inferring $N_{e}$ in the past is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation when searching for signals of selection. Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20. However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}. All sampling strategies lead to inflated ancient and deflated recent $N_{e}$ when neighborhood sizes were less than 20.

@@

In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes. How more sophisticated mixed-model methods would perform under our simulations is an interesting question that we plan to pursue in a future study, but statistical methods can only take us so far in the absence of controlled experiments. One currently popular approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}. We suggest a straightforward alternative for species in which the primary axes of population differentiation is space (note this is likely not the case for many modern human populations): run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, the structure corrections are insufficient. This is essentially the approach taken in our "clinal" model (though we also include normally distributed variation in our phenotypes). Of course it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.  \cjb{what do you think of the second half here?}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Apr 11 13:11:59 2019
\begin{table}[hbpt]
\small
\begin{adjustwidth}{-.9cm}{}
\centering
\caption{\bf Neighborhood size estimates from empirical studies.}
\begin{tabular}{rllrrll}
  \hline
 Species & Description & \makecell[l]{Neighborhood\\Size} & Method & Citation \\ 
  \hline
  \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60 - 37.80 & Genetic & \citep{Campbell1992}\\
  \makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14 - 27 & Genetic & \citep{Wright1943}\\ 
  \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20 - 30 & Genetic+Survey & \citep{Antlfinger1982} \\ 
  \makecell[l]{\textit{Oreamnos}\\\textit{americanus}} & mountain goat & 36 - 100 & Genetic & \citep{Shirk2014} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 40 - 213 & Genetic & \citep{Rousset1997}\\ 
  \makecell[l]{\textit{Formica sp.}} & colonial ants & 50 - 100 & Genetic & \citep{Pamilo1983} \\ 
  \makecell[l]{\textit{Astrocaryum}\\\textit{mexicanum}} & palm tree & 102 - 895 & Genetic+survey & \citep{Eguiarte1993} \\ 
  \makecell[l]{\textit{Spermophilus}\\\textit{mollis}} & ground squirrel & 204 - 480 & Genetic+Survey & \citep{Antolin2001}\\
  \makecell[l]{\textit{Sceloperus}\\\textit{olivaceus}} & lizard & 225 - 270 & Survey & \citep{Kerster1964} \\ 
  \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227 - 611 & Survey & \citep{Young1988} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & \citep{Rousset1997} \\ 
  \makecell[l]{\textit{Quercus}\\\textit{laevis}} & Oak tree & $>440$ & Genetic & \citep{Berg1995} \\ 
  \makecell[l]{\textit{Drosophila}\\\textit{pseudoobscura}} & fruit fly & 500 - 1,000 & Survey+Crosses & \citep{Wright1946} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{POPRES data\\NE Europe} & 1,342 - 5,425 & Genetic & \citep{Ringbauer2017} \\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
   \hline
\end{tabular}
\label{table:NStable}
\end{adjustwidth}
\end{table}

@@

Though our continuous space simulation allows incorporation of realistic demographic and spatial processes and is much faster than previous individual-based models, it is inevitably limited by the computational burden of tracking tens of thousands of individuals in every generation. In particular the calculations required for our mate selection and competition steps involve summarizing distances across all pairs of individuals and so scale very poorly \ak{how exactly does this scale? $O(N^2)$?} as the number of individuals within three-$\sigma$ increases. In part the issue of runtime scaling is a function of the spatial process itself -- under very limited dispersal we observed that coalescence requires over 30N generations, so forward-time methods must be run for a very long time to create a complete genealogy as underlies all real genome sequence data. The reverse-time model of continuous space evolution described in \cite{Barton2010} and implemented in \cite{Kelleher2014} may allow exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. However, incorporating selection or other processes into such models will likely be difficult.

>>>>

%Space is the Place: How Dispersal and Competition Shape Genetic Variation in Continuous Space
%Impacts of Continuous Spatial Structure on Analyses of Population Genetic Data
%Impacts of Isolation by Distance on Demographic Modeling and Association Studies
%Space is the Place: How Dispersal and Sampling Shape Inference From Genetic Variation in Continuous Space
%Evolution on the Final Frontier: How Dispersal and Sampling Shape Inference From Genetic Data in Continuous Space
\title{Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations. As a result many methods of analyzing genetic data assume that samples are a random draw from a randomly mating population, but are applied to real populations that are structured over space by limited dispersal. Here we implement a forward-time simulation of evolution in continuous space and use it to study the impacts of dispersal and sampling strategy on population genetic summary statistics, demographic inference, and genome-wide association studies. We find that continuous space skews the distribution of several common summary statistics, with the most bias occurring when Wright's neighborhood size is less than 100 and sampling is spatially clustered. In turn estimates of population size history based on the site frequency spectrum become biased in the face of limited dispersal. We show that in the context of genome-wide association studies the combination of spatial variation in the environment and limited dispersal creates spurious signals of genetic association with purely environmental phenotypes across much of the genome and that this bias is only partially corrected by regressing out principal components of ancestry. Last we discuss the relevance of our simulation results for inference from genetic variation in real organisms. 
\end{abstract}

@@

\section{Introduction}
The inescapable reality that biological organisms live, move, and reproduce in continuous spatial landscapes has been all but ignored in population genetic models. Indeed, a near universal rule of reproduction is that individuals mate with other nearby individuals, leading to a positive correlation between genetic and geographic distances. This pattern of "isolation by distance" \citep{Wright1943} is one of the most widely replicated empirical findings in population genetics \citep{Chen2017,Jay2012,Sharbel2000}, but classical mathematical models describing it are flawed approximations of the underlying process \citep{Felsenstein1975}.
As a result many studies focus on describing geographic structure as a set of populations connected by migration \citep{Malecot1948,Wright1931}, and analyze variation within clusters of genetic variation inferred by programs like \textit{STRUCTURE} \citep{Pritchard2000} assuming these are randomly mating units.

Though discrete models of between-population differentiation account for much of the genetic variation in species, the assumption that populations are randomly mating at some level has important implications for downstream inference of selection and demography. Methods based on the coalescent \citep{Kingman1982} assume that the sampled individuals are a random draw from a randomly mating population. If dispersal or mate selection is limited by space these assumptions are violated -- nearby individuals will be more closely related than an average random pair and drawing multiple samples from the same point on the landscape will represent a biased sample of the genetic variation present in the whole population. All downstream inferences drawn from the patterns of relationship inferred among sampled individuals will be subject to bias, but the extent depends on the degree of variation in ancestry across the landscape.

Our understanding of how many common summary statistics respond to variation in dispersal distance is limited. Tajima's $D$, for example, is often interpreted as reflecting evidence of selection or past population size changes (CITE \cjb{CJ: what's a good citation here?}). However this statistic is essentially a summary of the site frequency spectrum, which itself reflects variation in branch lengths of the underlying genealogy of sampled individuals. Because the spatial extent of dispersal and mate selection determines the rate of drift in a continuously distributed population \citep{Maruyama1972}, these population genetic summaries should also reflect variation in the scale of spatial interactions both over time and across populations. Similarly, the distribution of tract lengths of identity by state among individuals is often interpreted either through the lens of historic demography \citep{Harris2013} or selection \citep{Garud2015}. While that is so this distribution also contains abundant information about dispersal and mate selection \citep{Ringbauer2017,Baharian2016}. We are particularly keen on examining how such summaries will be affected by models that incorporate continuous space.

@@

The issue of fine-scale structure for putatively "within-population" analyses may have important implications for genome-wide association studies as well. This is because the genome-wide allele frequency variation caused by isolation by distance is similar to that expected when many sites in the genome are associated with a given phenotype \citep{Bulik-Sullivan2015}. When population structure and the environment covary across space genome-wide association studies (GWAS) may incorrectly infer phenotypic association at many SNPs that have differentiated by drift \citep{Mathieson2012}. This may be a particularly important effect when using GWAS to study selection on polygenic traits. Indeed two recent studies found that previous evidence of polygenic selection on human height in Europe was confounded by subtle population structure \citep{Sohail2018,Berg2018}, suggesting that existing methods to correct for population structure in GWAS are insufficient. However we have little quantitative idea of the population and environmental parameters that can be expected to lead to biases in GWAS. As the scale of sequence data now available for many species allows inference of increasingly fine-scale patterns of selection and demography, understanding how and when subtle spatial structure is likely to bias results is an important task for population genetics.

Here we describe an implementation of an individual-based model in continuous space that incorporates overlapping generations, a Gaussian dispersal kernel, and density-dependent fitness. The model scales to chromosome-scale alignments across tens of thousands of individuals and outputs the full genealogy and recombination history of all final-generation individuals. We use this simulation to test how sampling strategy interacts with isolation by distance to cause systematic variation in population genetic summary statistics typically analyzed assuming discrete population models. We then examine how the fine-scale spatial structures occurring under limited dispersal impact demographic inference from the site frequency spectrum. Lastly we examine the impacts of evolution in continuous space on genome-wide association studies (GWAS) and identify regions of parameter space under which the results from GWAS may be misleading.

\ak{I think we can cut the following paragraph if we just add the last sentence above} 
\cjb{agree - commenting out}
%Last we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations \citep{Price2006}. Because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within continuous populations can also deflate $p$ values in GWAS of traits influenced by the environment \citep{Mathieson2012}. Here we simulate both the underlying genealogy and phenotypes of a continuously distributed population and seek to identify regions of parameter space -- i.e. the strength of isolation by distance and the spatial distribution of environmental effects on phenotypes -- in which common methods employed for GWAS are subject to confounding of environmental and genetic effects on phenotypes.

@@

To isolate spatial effects from other components of the model such as overlapping generations, increased variance in reproductive success, and density-dependent fitness, we  also implemented a second version of the SLiM simulation in which mates are selected at random and offspring disperse to a random location on the landscape. We refer to these two models as the "spatial" and "random mating" models. For all simulations the full genealogy and recombination history of final-generation individuals were stored as tree sequences \citep{Kelleher2018}. Scripts for figures and analyses are available at https://github.com/petrelharp/spaceness.

@@

Following recent studies that showed strong signals for dispersal and demography in the distribution of shared haplotype block lengths \citep{Ringbauer2017,Baharian2016}, we also calculated various summaries of the distribution of pairwise identical-by-state (IBS) block lengths among samples. The full distribution of lengths of IBS tracts for each pair of individuals was first calculated with a custom python function (https://github.com/petrelharp/spaceness/blob/master/scripts/slimtools.py). We then calculated the first three moments of this distribution (mean, variance, and skew) and the number of blocks over $1e6$ base pairs both for each pair of individuals and for the full distribution across all pairwise comparisons.

@@

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of nongenetic phenotype variation for 1000 randomly sampled individuals in each spatial SLiM simulation and conducted a linear regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 0.5\% were excluded from this analysis. Phenotype values were set to vary by two standard deviations across the landscape in a rough approximation of the variation seen in height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model fully continuous spatial variation and compare GWAS output across a range of dispersal distances.

In the first simulation phenotypes for all individuals were drawn from a normal distribution with mean 100 and standard deviation 10.
\begin{equation}
    P \sim \mathcal{N}(100,10)
\end{equation}
Next we simulated clinal environmental influences on phenotype by drawing the phenotypes from independent normal distributions in which the mean was scaled by an individual's x position such that it varied by 2 standard deviations across the map. If $x$ is an individual's x coordinate on a 50x50 landscape, the phenotype is then:
\begin{equation}
    P \sim \mathcal{N}(100+\frac{2x}{5},10)
\end{equation}
Third, we approximate concentrated environmental effects by drawing phenotypes for individuals with x and y coordinates below 20 from a normal distribution with mean 2 standard deviations above the rest of the map. 
\begin{equation}
    P\sim
\begin{cases}
    \mathcal{N}(120,10),& \text{if } x< 20 \text{ and } y<20\\
    \mathcal{N}(100,10),& \text{otherwise}\\
\end{cases}
\end{equation}

@@

\begin{equation}
    P\sim
\begin{cases}
    \mathcal{N}(120,10),& \text{if } d < 3\\
    \mathcal{N}(100,10),& \text{otherwise}\\
\end{cases}
\end{equation}
where $d$ is the distance to the nearest patch.

@@

\subsection{Effects of Dispersal}
Limited dispersal inflates effective population size, creates correlations between genetic and spatial distances, and introduces subtle biases in the site frequency spectrum that are reflected in a positive Tajima's $D$ (Figure \ref{fig:sumstats}). At the extreme low end of dispersal distance this can result in an up to three-fold increase in genetic diversity relative to random-mating expectations. These effects are strongest when neighborhood sizes are below 100, but in combination with the effects of nonrandom sampling they can persist up to neighborhood sizes of at least 1000 (e.g. inflation in Tajima's $D$ and observed heterozygosity under midpoint sampling). Under random sampling the general pattern is similar to expectations of the original analytic model of \cite{Wright1943}, which predicts that populations with neighborhood sizes under 100 will differ substantially from random mating, while those above 10,000 will be nearly indistinguishable from panmixia.

@@

\subsection{Effects of Sampling}
One of the most important differences between random mating and spatial models is the effect of sampling: in a randomly mating population the spatial distribution of sampling effort has no effect on estimates of genetic variation, but when dispersal is limited sampling strategy can compound spatial patterns in the underlying genealogy and create pervasive impacts on all downstream genetic analyses. As expected we found that random sampling provides the most accurate summary of genetic diversity across the landscape. However this strategy is often impractical for empirical studies. In reality the difficulty of traveling through all parts of a species range and the inefficiency of collecting single individuals at each sampling site means that most studies follow something closer to the "point" sampling strategy we simulated, in which multiple individuals are sampled from nearby points on the landscape. For example, in ornithology a sample of 10 individuals per species per locality is a common target when collecting for natural history museums. In classical studies of \textit{Drosophila} variation the situation is considerably worse, in which a single orchard might be sampled with bated traps for instance.

@@

\subsection{Demography}
Classical population genetic models collapse many elements of life history variation into a single parameter, $N_{e}$, which is then taken to reflect the degree of variation present in the population when modeling the effects of selection or migration. Inferring $N_{e}$ in the past is now a common goal of population genomic analyses and an important step in establishing baseline expectations of genetic variation when searching for signals of selection. Here we found that one method of inference of historic $N_{e}$ based on genome-wide estimates of the site frequency spectrum, Stairwayplot \citep{Liu2015}, is relatively robust to variation in dispersal distance when sampling is random and neighborhood size is over 20. However, non-random sampling, and particularly midpoint sampling, causes the method to infer inflated estimates of past population sizes and a series of recent bottlenecks (Figure \ref{fig:demography}). All sampling strategies lead to inflated ancient and deflated recent $N_{e}$ when neighborhood sizes were less than 20.

@@

In summary, spatial covariation in population structure and the environment confound the interpretation of GWAS $p$-values, and correction using principal components is insufficient to fully separate these signals for polygenic traits under a variety of environmental and population parameter regimes. How more sophisticated mixed-model methods would perform under our simulations is an interesting question that we plan to pursue in a future study, but statistical methods can only take us so far in the absence of controlled environments. One currently popular approach to estimating the degree of bias in GWAS caused by population structure is LD score regression \citep{Bulik-Sullivan2015}. Though this approach appears to work well in practice, its interpretation is not always straightforward and it is likely biased by the presence of linked selection \citep{Berg2018}. We suggest a straightforward alternative for species in which the primary axes of population differentiation is space (note this is likely not the case for many modern human populations): run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations. If significant associations with sample locality are observed after correcting for population structure through PC regression or a kinship matrix, the structure corrections are insufficient. This is essentially the approach taken in our "clinal" model (though we also include normally distributed variation in our phenotypes). Of course it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.  \cjb{what do you think of the second half here?}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Apr 11 13:11:59 2019
\begin{table}[hbpt]
\small
\begin{adjustwidth}{-.9cm}{}
\centering
\caption{\bf Neighborhood size estimates from empirical studies.}
\begin{tabular}{rllrrll}
  \hline
 Species & Description & \makecell[l]{Neighborhood\\Size} & Method & Citation \\ 
  \hline
  \makecell[l]{\textit{Ipomopsis}\\\textit{aggregata}} & flowering plant & 12.60 - 37.80 & Genetic & \citep{Campbell1992}\\
  %\makecell[l]{\textit{Linanthus}\\\textit{parryae}} & flowering plant & 14 - 27 & Genetic & \citep{Wright1943}\\ %this one got debunked by a series of studies in the 90's that found that spatial differentiation in flower color is driven by environmental variability across years. See https://doi.org/10.1111/j.1558-5646.2007.00219.x
  \makecell[l]{\textit{Borrichia}\\\textit{frutescens}} & salt marsh plant & 20 - 30 & Genetic+Survey & \citep{Antlfinger1982} \\ 
  \makecell[l]{\textit{Oreamnos}\\\textit{americanus}} & mountain goat & 36 - 100 & Genetic & \citep{Shirk2014} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 40 - 213 & Genetic & \citep{Rousset1997}\\ 
  \makecell[l]{\textit{Formica sp.}} & colonial ants & 50 - 100 & Genetic & \citep{Pamilo1983} \\ 
  \makecell[l]{\textit{Astrocaryum}\\\textit{mexicanum}} & palm tree & 102 - 895 & Genetic+survey & \citep{Eguiarte1993} \\ 
  \makecell[l]{\textit{Spermophilus}\\\textit{mollis}} & ground squirrel & 204 - 480 & Genetic+Survey & \citep{Antolin2001}\\
  \makecell[l]{\textit{Sceloperus}\\\textit{olivaceus}} & lizard & 225 - 270 & Survey & \citep{Kerster1964} \\ 
  \makecell[l]{\textit{Dieffenbachia}\\\textit{longispatha}} & \makecell[l]{beetle-pollinated\\colonial herb} & 227 - 611 & Survey & \citep{Young1988} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{Gainj- and Kalam- \\speaking people,\\Papua New Guinea} & 410 & Survey & \citep{Rousset1997} \\ 
  \makecell[l]{\textit{Quercus}\\\textit{laevis}} & Oak tree & $>440$ & Genetic & \citep{Berg1995} \\ 
  \makecell[l]{\textit{Drosophila}\\\textit{pseudoobscura}} & fruit fly & 500 - 1,000 & Survey+Crosses & \citep{Wright1946} \\ 
  \makecell[l]{\textit{Homo sapiens}} & \makecell[l]{POPRES data\\NE Europe} & 1,342 - 5,425 & Genetic & \citep{Ringbauer2017} \\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 240,000 & Survey & \citep{Rousset1997}\\ 
  \makecell[l]{\textit{Bebicium}\\\textit{vittatum}} & intertidal snail & 360,000 & Genetic & \citep{Rousset1997}\\ 
   \hline
\end{tabular}
\label{table:NStable}
\end{adjustwidth}
\end{table}

@@

One complication in the inference of any spatial demographic parameter is the balance between local and global process. Many species are structured locally by limited dispersal, but also contain deeply divergent lineages in different regions that reflect signals of ancient episodes of geographic isolation or strong barriers to dispersal. Gene flow upon secondary contact of two previously isolated lineages should create clinal patterns similar to isolation by distance, and it will be difficult to determine when inferred dispersal parameters are reflecting recent demographic process versus the historic patterns of geographic isolation. In addition, spatially varying selection will create allele frequency variation over space that may mimic isolation by distance. Indeed, a series of field studies described in \cite{Schemske2007} found that in Wright's original empirical example of isolation by distance, the flowering plant \textit{Linanthus parryae}, patterns of flower color differentiation over space primarily reflect temporal and spatial variation in selection rather than limited dispersal. Studies simulating selection and dispersal interacting in space (e.g. \cite{Ralph2010}) and testing for identifiability of inferred dispersal or selective parameters may offer new insight into the extent of our ability to accurately infer evolutionary processes in real systems.

%Recent methods that aim to estimate geographic barriers to dispersal \citep{Ringbauer2018,Petkova2015} partially address this issue by attempting to infer the location of regions resistant to gene flow, but here again the balance between historical and current demographic processes represented in our inference is unclear. Limiting analyses to low-frequency alleles \citep{Novembre2009} or between long haplotype blocks \citep{Ringbauer2017} is one approach to limiting inference to recent demography (because these genomic features should arise primarily from patterns in the most recent branches of the genealogy), but as we showed spatial processes also cause shifts in branch lengths throughout the genealogy. Further studies on the ways in which the sample genealogy responds to variation in dispersal both over time and across space are a good direction for future research.

Though our continuous space simulation allows incorporation of realistic demographic and spatial processes and is much faster than previous individual-based models, it is inevitably limited by the computational burden of tracking tens of thousands of individuals in every generation. In particular the calculations required for our mate selection and competition steps involve summarizing distances across all pairs of individuals and so scale very poorly \ak{how exactly does this scale? $O(N^2)$?} as the number of individuals within a three-$\sigma$ radius increases. In part the issue of runtime scaling is a function of the spatial process itself -- under very limited dispersal we observed that coalescence requires over 30N generations, so forward-time methods must be run for a very long time to create a complete genealogy as underlies all real genome sequence data. The reverse-time model of continuous space evolution described in \cite{Barton2010} and implemented in \cite{Kelleher2014} may allow exploration of parameter regimes with population and landscape sizes more directly comparable to empirical cases like humans. However, incorporating selection or other processes into such models will be difficult.
